---
title: 'STAT 413/613 HW 2: stringr and lubridate'
author: "Jaehee Lee"
date: "`r Sys.Date()`"
output:
  html_document: 
    number_sections: true
editor_options:
  chunk_output_type: inline
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align  = "center",
                      fig.height = 3, 
                      fig.width  = 5)
```

# Instructions {-}

Rename the starter file under the analysis directory as `hw_02_yourname.Rmd` and use it for your solutions.   
1. Modify the "author" field in the YAML header.  
2. Stage and Commit R Markdown and HTML files (no PDF files).   
3. **Push both .Rmd and HTML files to GitHub**.   
- Make sure you have knitted to HTML prior to staging, committing, and pushing your final submission.  
4. **Commit each time you answer a part of question, e.g. 1.1**   
5. **Push to GitHub after each major question**, e.g., Scrabble and Civil War Battles   
- **Committing and Pushing are graded elements for this homework.**  
6. When complete, submit a response in Canvas  
    
- Only include necessary code to answer the questions.
- Most of the functions you use should be from the tidyverse. Too much base R will result in point deductions.
- Use Pull requests and or email to ask me any questions. If you email, please ensure your most recent code is pushed to GitHub.  

- Learning objectives:
    - Manipulate dates and times with lubridate.


# Scrabble Words

For this exercise, we are using the [Collins Scrabble Words](https://en.wikipedia.org/wiki/Collins_Scrabble_Words), 
which is most commonly used outside of the United States. The dictionary most often used in the United States is the 
[Tournament Word List](https://en.wikipedia.org/wiki/Official_Tournament_and_Club_Word_List).

**WARNING**: Do not try `str_view()` or `str_view_all()` on these data.It will stall your computer.
load packages
```{r}
library(tidyverse)
library(stringr)
library(readr)
library(lubridate)
```

1. Use a readr function to load the 2015 list of Collins Scrabble Words into R from your data folder or from 
   <https://data-science-master.github.io/lectures/data/words.txt>  
   + (note: "NA" is an official Scrabble word).
```{r}
read_delim('https://data-science-master.github.io/lectures/data/words.txt', delim = "|") -> collins_scrabble
```

2. What are the six longest words that have the most "X"'s in them?

```{r}

collins_scrabble %>%
  mutate(length_of_word = str_length(word)) %>%  #make new column for length_of_word
  mutate(count_X = str_count(word,"X")) %>% #make new column for counting "X"
  arrange(desc(count_X)) -> collins_scrabble_1 
  
collins_scrabble_1 %>%  
  select(count_X) %>%  
  head(1) -> most_X  #This will tell me the most "X"'s count which is 2

collins_scrabble_1 %>%  
  filter(count_X == most_X$count_X[1]) %>% 
  arrange(desc(length_of_word)) %>%
  head() %>%  # show 6 longest words 
  select(word)
```


3. How many words have an identical first and second half of the word? If a word has an odd number of letters, exclude the middle character.  
  - MURMUR counts because MUR is both the first and second half.
  - JIGAJIG counts because the middle A is excluded so JIG is both the first and second half. 
  - Save the results to a variable.
```{r}
collins_scrabble_1 %>%  
  mutate(is_even = if_else(str_length(word) %% 2 == 0,  "yes", "no")) %>% 
  mutate(first_half = if_else(is_even == "yes", 
                              str_sub(word, 1, length_of_word/2), str_sub(word, 1, floor(length_of_word/2)))) %>%
  mutate(second_half = if_else(is_even == "no", 
                               str_sub(word, (length_of_word/2)+1, length_of_word), str_sub(word, ceiling(length_of_word/2 +1), length_of_word))) %>%  
  filter(first_half == second_half) -> collins_scrabble_2

collins_scrabble_2 %>% 
  tally() -> number_identical_first_and_second_half
```

  
4. Use the results from 3 to find the longest word with an identical first and second half of the word?

```{r}
collins_scrabble_2 %>% 
  arrange(desc(length_of_word)) %>%  
  head(1) %>% 
  select(word)
  
```

# Civil War Battles

The data in "civil_war_theater.csv" contains a information on American Civil War battles, taken from [Wikipedia](https://en.wikipedia.org/wiki/List_of_American_Civil_War_battles).  

Variables include:   

- `Battle`: The name of the battle.
- `Date`: The date(s) of the battle in different formats depending upon the length of the battle. 
  + If it took place on one day, the format is "month day, year". 
  + If it took place over multiple days, the format is "month day_start-day_end, year". 
  + If it took place over multiple days and months, the format is "month_start day_start - month_end day_end, year". 
  + If it took place over multiple days,months, and years, the format is "month_start day_start, year_start - month_end day_end, year_end".
- `State`: The state where the battle took place. Annotations (e.g.     describing that the state was a territory at the time) are in parentheses.
- `CWSAC`: A rating of the military significance of the battle by the Civil War Sites Advisory Commission. `A` = Decisive, `B` = Major, `C` = Formative, `D` = Limited.
- `Outcome`: Usually `"Confederate victory"`, `"Union victory"`, or `"Inconclusive"`, followed by notes.
- `Theater`: An attempt to to identify which theater of war is most associated with the battle
    
1. Use a readr function and relative paths to load the data into R.
```{r}
civil_war_data <- read_csv("civil_war_theater.csv")
str(civil_war_data)
```

The next several questions will help you take the dates from all the different formats and create a consistent set of start date and end date variables in the data frame. We will start by calculating how many years, and months are in each battle.

2. Add a variable to the data frame with the number of years for each battle.  

- Create a character variable as follows. This can be used as a pattern in a regular expression.  
   
- Use `year_regex` to now count the number of years in each battle, add this to the data frame,  and save the data frame.
```{r}
year_regex <- stringr::str_c(1861:1865, collapse = "|")
year_regex

civil_war_data %>%  
  mutate(year = str_extract_all(Date, year_regex)) %>%  
  mutate(year_count = str_count(Date, year_regex))->civil_war_data
```

3. Add a variable to the data frame with the number of months for each battle.  
- Consider R's built-in vector of month names: `month.name`.

- Use `month.name` to count the number of month names in the `Date` variable in each battle. 
- Add this to the data frame. (You might need to do something similar to what we did in part 2).
```{r}
month_regex <- stringr::str_c(month.name, collapse = "|") 
month_regex
civil_war_data %>%  
  mutate(month_names = str_extract_all(Date, month_regex)) %>%  
  mutate(month_count = str_count(Date, month_regex)) -> civil_war_data
```


4. Add a variable to the data frame that is `TRUE` if `Date` spans multiple days and is `FALSE` otherwise. Spanning multiple months and/or years also counts as `TRUE`.
```{r}
civil_war_data %>%  
  mutate(hyphen = str_detect(Date, "-")) %>%  
  mutate(multiple_days = if_else(hyphen == "TRUE",  "TRUE", "FALSE")) -> civil_war_data

```


5. Make four new data frames by filtering the data based on the length of the battles:  
   + a data frame with the data for only those battles spanning just one day, 
```{r}
civil_war_data %>%
  filter(multiple_days == "FALSE") -> one_day
```
   
   + a data frame with the data for only those battles spanning multiple days in just one month, 
```{r}
civil_war_data %>%  
  filter(multiple_days == "TRUE") %>%  
  filter(year_count == 1) %>%  
  filter(month_count == 1) %>%  
  filter(hyphen == "TRUE") -> multiple_days_one_month
```
   
   
   + a data frame with the data for only those battles spanning multiple months but not multiple years, and,
```{r}
civil_war_data %>%  
  filter(multiple_days == "TRUE") %>%  
  filter(year_count == 1) %>%  
  filter(month_count == 2) -> multiple_months_not_years
```
   
   + a data frame with the data for only those battles spanning multiple years.
```{r}
civil_war_data %>% 
  filter(multiple_days == "TRUE") %>%  
  filter(year_count == 2) -> multiple_years
```
   


6. For each of the four new data frames,  
  - Add two new variables:  
    + `Start` should contain the start-date of each battle. 
    + `End` should contain the end-date of each battle.
    + Hint: look at help for `separate()`
    + Make sure these are `Date` class objects. 
  - Remove the `Date` variable from  each data frame.
  - Save the data frames with the new variables
```{r}
# 1 
one_day %>% 
  mutate(Start = Date) %>%  
  mutate(End = Date) %>%  
  select(-Date)-> one_day_updated
#2 
separate(data = multiple_days_one_month, col = Date, sep = "-", into = c("Start", "End")) -> multiple_days_one_month_updated
multiple_days_one_month_updated %>%  
  unite(Start, c("Start", "year")) %>%  
  unite(End, c(End, month_names)) -> multiple_days_one_month_updated
# 3 
separate(data = multiple_months_not_years, col = Date, sep = "-", into = c("Start", "End")) -> multiple_months_not_years_updated
multiple_months_not_years_updated %>%  
  unite(Start, c("Start", "year")) -> multiple_months_not_years_updated

# 4
separate(data = multiple_years, col = Date, sep = "-", into=c("Start", "End")) -> multiple_years_updated
```
  
7. Create a new data frame with all the battles and the Start and End  dates by binding the rows of the four data frames as updated in part 6 

8. Calculate the number of days each battle spanned.  
- What's the longest battle of the war? 
- How long did it last?

9. Is there an association between the CWSAC significance of a battle and its duration?  
- Create an appropriate plot. 
- Interpret the plot in one sentence to answer the question.
- Extra Credit: Test for a linear relationship using `lm()` and interpret the results in one sentence based on the $p$-value and adjusted R-squared.
    
10. Extra Credit: Did the [theaters of war](https://en.wikipedia.org/wiki/Theater_(warfare)) shift during the American Civil War?  
- Reproduce this plot in R (filter out states with two or fewer battles.)
- Interpret the plot in one sentence.




















